---
title: "EDA of Sacramento House Prices"
author: "Lesley Miller"
date: "2/8/2020"
output: 
   github_document:
      toc: true
bibliography: citations.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
```

```{r load libraries, echo=FALSE}
# load packages
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(mice))
suppressPackageStartupMessages(library(ggcorrplot))
library(knitr)
library(patchwork)
```


```{r load data, tidy=TRUE, echo=FALSE, include=FALSE}
# load training  data 
df <- read_csv("data/X_train.csv")
#df_y <- read_csv("data/y_train.csv")
```

# Data Description 
[Data Source](https://support.spatialkey.com/spatialkey-sample-csv-data)

> This dataset contains information on various attributes of homes sold in and around the Sacramento, California. The data is from real estate transactions taking place over a 5 day period. The data is from the website for the `SpatialKey` software and can be loaded using the R caret package [@caret] or downloaded as a `csv` from [here](https://support.spatialkey.com/spatialkey-sample-csv-data). 

```{r echo=FALSE}
num_examples <- nrow(df)
num_features <- ncol(df)
print(paste("The number of training examples is:", num_examples))
print(paste("The number of predictors is:", num_features))
```

## Missing Data
> The graphic below shows that all 521 rows of the dataset are complete. 

```{r echo=FALSE}
md.pattern(df, plot = FALSE)
```



## Summary of Predictors 
|Predictor | Type | Description |
|-----------|:--------:|---------:|
|city | categorical| The city where the house is located.|
|zip | categorical | The zip code where the house is located.|
|beds | categorical | The number of bedrooms in the house.|
|baths | categorical | The number of bathrooms in the house.|
|type | categorical | The type of dwelling as either `Residential`, `Condo` or `Multi-Family`. |
|sqft | continuous | The size of the house measured in square feet.|
|latitude| continuous | The latitude geospatial location of the house.|
|longitude| continuous | The longitude geospatial location of the house.|

## Distribution of Categorical Predictors 

> The most common number of bedrooms for a house is 3 with the second most common number being 4 bedrooms. Just under 100 homes of the 521 in the training set contain 2 bedrooms and very few homes contain 1, 5 or 6 bedrooms. 

> The majority of the houses (~ 2/3) contain 2 bathrooms. One bathroom was second most common. Having 3 bathrooms was only present in about 15% of homes and having 1.5, 2.5, 3.5, 4, 4.5 or 5 bathrooms was very rare. 

> Almost all of the homes were classifed as `Residential` with extremely few homes being classifed as `Condo` or `Multi-family`. 


```{r echo=FALSE}
# convert zipcodes to numeric vector  by removing z char
# remove z character from zip codes 
df <- df %>% 
   mutate(
   zip = as.numeric(gsub(pattern = "z", replacement = "", x = df$zip))
)

# get the categorical variables 
cat_vars <- df %>% 
      select(-sqft, -latitude, -longitude)

# get the numeric variables 
numeric_vars <- df %>% 
   select(-city, -type)
```

```{r echo=FALSE}
# make a list of frequency tables 
tables <- list()
for (i in 1:length(cat_vars)) {
      freq_table <- as_tibble(as.data.frame(table(cat_vars[[i]])))
      col_name <- colnames(cat_vars)[i]
      tables[[col_name]] <- freq_table
}

```


```{r echo=FALSE}
plots <- list()
# plot the frequency tables of beds, baths & type 
for (i in 3:length(tables)) {
   table <- tables[[i]]
   p <- table %>% 
      ggplot(aes(x = reorder(Var1, -Freq), y = Freq, color = Var1)) + 
      geom_col() + 
      labs(title = paste("Distribution of", names(tables)[i]),
           x = names(tables)[i],
           y = "Frequency") + 
      theme_grey() + 
      theme(legend.position = "none")
   plots[[names(tables)[i]]] <- p
}

# display the plots 
wrap_plots(plots, nrow = 3)
```

## Distribution of House Size
```{r echo=FALSE}
# make tibble of range of house size 
house_size_range <- tibble(`Smallest House Size` = range(df$sqft)[1],
                           `Largest House Size` = range(df$sqft)[2])
kable(house_size_range, caption = "Range of House Square Footage")
```

> The distribution of house size is pulled to the right with outliers of large house size. The homes range in size from 484 sq ft on the low end to 4878 sq ft on the high end. The peak of the density centers around ~ 1266 sq ft. 

```{r continuous vars, echo=FALSE}

# get the peak house size
peak_house_size <- density(df$sqft)$x[119]

# make a density plot of the house square footage
df %>% 
   ggplot(aes(x = sqft)) + 
   geom_density(color = "turquoise") + 
   labs(title = "Distribution of House Size",
        x = "Square Footage",
        y = "Density") + 
   theme_grey() + 
   geom_vline(xintercept = peak_house_size) +
   annotate(geom = "text", 
            x = 1800, 
            y = 1.5e-04, 
            label = paste("Peak House Size\n", 
                          round(peak_house_size,1), "sqft"))
```

> As expected, with a rising number of bedrooms, the bulk of the distribution for house size also rises but there is variability within different levels of bedrooms. After one-bedrooms, there is a wide range of sizes for homes with 2 bedrooms and beyond. However, as the number of bedrooms rises, the density becomes thinner and thinner as fewer and fewer homes are have 4 or more bedrooms. 

```{r echo=FALSE}
df %>% 
   ggplot(aes(x = as.factor(beds), y = sqft, color = as.factor(beds))) +
   geom_violin() +
   theme_grey() + 
   theme(legend.position = "none") + 
   labs(title = "Distribution of square footage by number of bedrooms", 
        x = "Number of Bedrooms",
        y = "House Size (sqft)") 
```

> Like with number of bedrooms, the house size increases with the number of bathrooms present but there still is considerable variability. Houses with 1.5 bathrooms have a very small range of house sizes. The density of the distributions becomes very thin as the number of bathrooms increases. Something potentially to consider is since there is considerable variability and overlap in size for houses with different bathrooms the question becomes if number of bathrooms will be a good predictor. On the extreme ends of bath number, the distributions of size are different enough and do not overlap.

```{r echo=FALSE}
df %>% 
   ggplot(aes(x = as.factor(baths), y = sqft, color = as.factor(baths))) +
   geom_violin() + 
   theme_grey() + 
   theme(legend.position = "none") + 
   labs(title = "Distribution of square footage by number of bathrooms", 
        x = "Number of Bathrooms",
        y = "House Size (sqft)")

```

# Correlation Matrix 
>

```{r correlation matrix, echo=FALSE}
corr <- cor(numeric_vars)
ggcorrplot(corr, 
           method = "square",
           type = "lower", 
           outline.col = "white",
           ggtheme = "theme_grey",
           lab = TRUE,
           colors = c("green", "white", "magenta"),
           title = "Correlation Matrix of all numeric predictors") 
```



# Conclusions
### Potential Problems with the data 
> After conducting this visual exploration, there doesn't seem to be any problems to report. The dataset is complete and the variables have reasonable distributions (i.e. nothing is terribly skewed)

### Is the data appropriate to the question 
> The research question for this project will be about discovering the relation between house price and the predictors. How will changing the values of the predictors influence the house price? Are there predictors that affect the house price more than others? This dataset provides some basic attributes about a house that should prove to be useful in inferring house price (i.e. housing type, bed and bath number etc.)

### Research question formulation 
> The primary inferential research question for this probject will be`

- How are the predictors related to the response? (linear? polynomial? etc)

> Related questions will include:

- Which of the predictors are most strongly associated with the response of house price?

- If the bed and bathroom number or house size increase, how does this affect the house price?

#### Analysis Packages
The tools for this exploratory analysis include the R language [@R], and the following R packages: tidyverse [@tidyverse], mice [@mice], knitr [@knitr], here [@here], ggcorrplot [@ggcorrplot] and patchwork [@patchwork].

This analysis uses the `Sacramento` dataset obtained through the R caret package [@caret].

# References


